Volatility and Correlations --- Eyes that the wise have in their head

### 2014-09-29 thesis/SLV_Unconditional.tex
Add the draft of the paper.

\chapter{Return Models}\label{chp:PriceModels}
In this chapter we review some of the discrete-time return models and
fit them to intraday returns. The intention is to assess the
applicability of these models and to understand their statistical
properties. The implications of these properties on the returns'
correlation and covariance matrix are studied in chapters
\ref{chp:Gaussian} and \ref{chp:CrossCorrelationFat}.

\S\ref{sec:Garch_model} reviews the GARCH model using the Nordea
15-minute returns as an example. \S\ref{sec:SLV_model} examines the
stochastic log-volatility model and discusses the possibility of
describing the residuals of the log-volotility series using the
Johnson Su distribution. The same Nordea return series is fitted as an
example. \S\ref{sec:XieCalc} derives the unconditional distribution
function of the stochastic log-volotility model and compares the
analytic results to the 30-minute returns of a number of Swedish
stocks.

\section{Gaussian Distribution}
The justification of modeling return series as independent,
identically distributed Gaussian variates comes from imagining the
price process $S(t)$ as a Brownian motion whose increment $\sigma
dw_t$ at each time step is independent and scales as $\sigma
\sqrt{dt}$. Then, by adding a drift term $\mu dt$ that represents some
deterministic trend in the price process, one can express the
price $S(t)$ as a stochastic differential equation:
\begin{eqnarray*}
  dS &=& S\mu dt + S\sigma dw_t \\
  d(\ln S) &=& (\mu - \frac{1}{2} \sigma^2)dt + \sigma dw_t \\
\end{eqnarray*}
It follows from this equation that $\ln S(t) - \ln S_0$, where $S_0$
is the price at time 0, has Gaussian distribution with mean $(\mu -
\sigma^2/2)t$ and variance $\sigma^2 t$. Therefore, in a discrete-time
model, where the length of each time step $\Delta t$ is fixed, the
return $r_t = \ln S(t + \Delta t) - \ln S(t)$ is assumed to be
Gaussian distributed and have mean and variance that are functions of
$\Delta t$.

The picture depicted above is of course overly simplified, and as we discussed in
\S\ref{sec:StylizedFacts}, the distribution of returns is not really
Gaussian. Nevertheless, the assumption of Gaussian distributed returns
underlies such important theories as Black and Scholes theory of
option pricing. Thus, albeit hugely inaccurate, the Gaussian
distribution as a returns model shall not be forgotten. In the next
few sections, we discuss some more realistic returns models.

\section{GARCH models} \label{sec:Garch_model}
``GARCH'' is the acronym for ``Generalized Autoregressive Conditional
Heteroscedasticity''. A GARCH(p, q) model is defined by the following
equation system:
\begin{equation}
  \label{eq:garch_def}
  \begin{aligned}
    r_t &= \mu + \epsilon_t \\
    \sigma_t^2 &= \alpha_0 + \sum_{i=1}^q \alpha_i \epsilon_{t-i}^2 +
    \sum_{i=1}^p \beta_i \sigma_{t-i}^2
  \end{aligned}
\end{equation}
where $\epsilon_t$ is a random variable of a prescribed distribution,
the most common ones being Gaussian, Student's t, Asymmetric Student's
t, etc; $\sigma_t^2$ is the variance of $\epsilon_t$
\cite{Bollerslev86}. The autocorrelation function of $\epsilon_t^2$ is
given by (c.f. \cite{Bollerslev87})
\begin{eqnarray*}
\rho_n &=& \sum_{i=1} ^{\max(p, q)} (\alpha_i + \beta_i) \rho_{n-i}
\text{ for $n > p$}
\end{eqnarray*}
where $\alpha_i$ with $i > q$ and $\beta_i$ with $i > p$ are taken as
zeros. It is clear from this equation that the partial autocorrelation
function cuts off at $\max(p, q)$.

To assess the applicability of GARCH models at the face of intra-day
returns, we study the {\it Nordea Bank 15min} returns sampled during
the period 2012/01/16 - 2012/04/20. Firstly we plot the autocorrelation
function (ACF) of $\epsilon_t^2$ from this sample, as shown
in figure \ref{fig:nordea_15min_acf}.

At the absence of GARCH effects, the ACF is expected to have an
asymptotic Gaussian distribution with mean 0 and variance 1/T
\cite{Bollerslev86, Bollerslev87}, which is clearly not the case in
figure \ref{fig:nordea_15min_acf}: The first 5 autocorrelations have
comparable sizes and do not fall off as in a Gaussian
scheme. Moreover, figure \ref{fig:nordea_15min_vlt_acf} shows 
even more clearly that the conditional variances of the series are
correlated. These observations suggest a GARCH(p, q) model can be
appropriate.

\begin{figure}[htb!]
  \centering
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_acf.pdf}
    \label{fig:nordea_15min_acf}
  }
  \subfigure[]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_vlt_acf.pdf}
    \label{fig:nordea_15min_vlt_acf}
  }
  \caption{\small \it Auto-correlations among Nordea Bank 15min
    returns. \ref{fig:nordea_15min_acf}: Auto-correlations (ACF) of the
    squared returns; \ref{fig:nordea_15min_vlt_acf}: Auto-correlations
    (ACF) of the conditional variances.}
\end{figure}

According to Andersen and Bollerslev et al \cite{Andersen03},
$\epsilon_t$ can be decomposed as $\epsilon_t = \sigma_t z_t$, where
$z_t \sim N(0, 1)$ and $\sigma_t$ can be estimated using realized
volatility. To find out whether this is true for the aforementioned
series we look at the probability plot of $\epsilon_t/\sigma_t$
(figure \ref{fig:nordea_bank_15min_z_prob}), which is consistent with
a standard Gaussian distribution except for a few outliers at the two
ends. Hence the fat tails of the probability density function can be
entirely accounted for by GARCH effects, implying a Gaussian
conditional distribution is sufficient for $z_t$.
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=80 258 104
    220]{../pics/nordea_bank_15min_z_prob.pdf}
  \caption{\small \it{Probability plot of $z_t = \epsilon_t /
      \sigma_t$. $\epsilon_t$ are derived from Nordea Bank 15min
      returns while $\sigma_t$ are realized volatilities calculated
      using 30s returns within each 15min interval. Horizontal axis:
      $z_t$}; Vertical axis: cummulative probability function (CDF) of
    $z_t$, arranged on such a scale that the CDF of the standard
    Gaussian is a straight line.}
  \label{fig:nordea_bank_15min_z_prob}
\end{figure}

To determine the orders p and q of the GARCH model, we again check
the partial autocorrelation function of $\epsilon_t^2$. As just
mentioned, the first 5 partial autocorrelations are significant. Thus
we can tentatively estimate a GARCH(1, 5) model. Then by applying the
Lagrange multiplier test and accordingly removing insignificant
parameters, we obtain a GARCH(1, 3) model
\ref{tab:nordea_15min_garch}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    parameter & value & std. error & t statistic \\
    \hline
     $\alpha_0$ &   1.05668e-06 &   3.56871e-07 &    2.96096\\
     \hline
     $\beta_1$  &     0.687212  &   0.0299314   &    22.9596\\
     \hline
     $\alpha_1$ &      0.128614 &   0.0268144   &    4.79643\\
     \hline
     $\alpha_3$ &    0.0568022  &   0.0301503   &    1.88397 \\
     \hline
  \end{tabular}
  \caption{\small \it GARCH(1, 3) model of Nordea Bank 15min returns}
  \label{tab:nordea_15min_garch}
\end{table}

\section{ARIMA Log-volotility Models}\label{sec:SLV_model}
With the availability of transaction data, estimating conditional
volatility using returns sampled at a higher frequency (realized
volatility) gives superior acuracy and reliability. However, at which
frequency the time series should be sampled in order to give an
unbiased and consistent estimate of the volatility is not a trivial
question. Naively one would argue by the doctrine of statistics that
the often the series is sampled, the better the estimate. This is also
well supported in theory, since the quadratic variation which is the
theoretial variance of the time series (see \cite{Andersen03}, theorem
1 and corollary 1), is approximated as (see \cite{Protter05}, ch.II,
\S6, theorem 22):
\[
\var(p_{t+h} - p_t) = [r, r]_{t+h} - [r, r]_t = \lim_{n \to \infty}
\sum_{T^i \in \pi_n} (p^{T_{i+1}} - p^{T_i})^2
\]
where $p_t$ is the log price process and $r_t = p_t - p_0$. $\pi_n$
is a refining partition of the interval $[t, t+h]$.

However, due to the noise introduced by market micro-structure, there
is actually an optimal sampling frequency relative to $h$. While the
method to determine this optimal frequency is a subject of debate (see
for example \cite{Sahalia05}), it is not difficult to find a fairly
satisfactory frequency in practice.

Andersen and Bollerslev et al proved in \cite{Andersen03} that, with
the conditional volatility well approximated by the realized
volatility, $r_{t+h, h} = p_{t+h} - p_t$ is governed by a Gaussian
distribution (see theorem 2 of \cite{Andersen03}):
\[
r_{t+h, h} | \sigma(\mu_{t+s}, \sigma_{t+s})_{s \in [0, h]} \sim
N(\int_{0}^h \mu_{t+s} ds, \int_{0}^h \sigma_{t+s}^2 ds)
\]

Thus one can try a few frequencies and compare the distribution of
$r_{t+h, h}/\hat{\sigma}_{t+h, h}$ with the standard Gaussian. If the
two match, $\hat{\sigma}_{t+h, h}$ is a good estimate of $\sigma_{t+h,
  h}$, the volatility of $r_{t+h, h}$.

For the Nordea Bank 15-minute returns, $h = 15\text{min}$, and a
reasonally good sampling frequency is 1/30sec, as confirmed by the
probability plot \ref{fig:nordea_bank_15min_z_prob} as well as by the
autocorrelation functions of $r_{t+h, h} / \sigma_{t+h, h}$ and of
$(r_{t+h, h} / \sigma_{t+h, h})^2$ which are shown in
\ref{fig:nordea_15min_quotient_acf} and
\ref{fig:nordea_15min_quotient_squared_acf} respectively.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $r_{t+h, h} / \sigma_{t+h, h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_quotient_acf.pdf}
    \label{fig:nordea_15min_quotient_acf}
  }
  \subfigure[ACF of $(r_{t+h, h} / \sigma_{t+h, h})^2$]{
    \includegraphics[scale=0.4, clip=true, trim=95 236 118
    200]{../pics/nordea_15min_quotient_squared_acf.pdf}
    \label{fig:nordea_15min_quotient_squared_acf}
  }
  \caption{\small \it Nordea 15min $r_t/\sigma_t$ and
    $(r_t/\sigma_t)^2$ ACF.}
\end{figure}

We see in figure \ref{fig:nordea_15min_quotient_acf} and
\ref{fig:nordea_15min_quotient_squared_acf} there is essentially no
autocorrelation in the series $r_{t+h, h} / \sigma_{t+h, h}$ and
$(r_{t+h, h} / \sigma_{t+h, h})^2$, supporting the assumption that the
series is i.i.d Gaussian distributed. As a result, the problem of
modeling $r_{t+h, h}$ boils down to modeling $\sigma_{t+h,
  h}$. Andersen and Bollerslev et al reported that, for the exchange
rates between Deutch mark, yen and dollar, $\ln \sigma_{t+h, h}$ is
gaussian distributed \cite{Andersen03}. This is, however, not the case
for our series in question. In fact, in our case, $\ln \sigma_{t+h,
  h}$ is right skewed (skewness 0.3342) and leptokurtic (kurtosis
6.1006). See figure \ref{fig:nordea_15min_logvol_prob}
\begin{figure}[htb!]
  \centering
  %\vspace{-15mm}
  \includegraphics[scale=0.5, clip=true, trim=80 223 107
  4]{../pics/nordea_15min_logvol_prob.pdf}
  \caption{\small \it Probability plot of Nordea 15min
    $\ln\sigma_{t+h, h}$ unconditional distribution}
  \label{fig:nordea_15min_logvol_prob}
\end{figure}

Moreover, the series of $\ln\sigma_{t+h, h}$ shows long-lasting and
periodic autocorrelations with an apparent period of 33 (see figure
\ref{fig:nordea_15min_logvol_acf}). This suggests the series may be
described by an integrated seasonal ARMA\footnote{ARMA ---
  Autoregressive Moving Average} model. But before we set out
to specify and estimate this model of the logvolatility
$\ln\sigma_{t+h, h}$, it is useful to first write down the model of
the returns $r_t$:
\begin{eqnarray}
   r_{t+h, h} &=& \ln p_{t+h} - \ln p_t \nonumber \\
   r_{t+h, h} &=& \mu + \sigma_{t+h, h} b_t \label{eq:SLV_spec}
\end{eqnarray}
where $p_t$ is the price of the asset at time $t$ and $b_t \sim N(0,
1)$. For simplicity, in the rest of this chapter we shall just write
$t$ for the subscript ``t+h, h'', since the time interval $h$ is fixed
for each time series and is a known constant. In the case of the
Nordea Bank 15-minute returns, h = 15 minutes.

As has just been said, $\ln \sigma_t$ for the series in question has
long-lasting auto-correlations (often referred to as ``long
memory''). This suggests simplifying the series by differencing
\cite{BoxJenkins94}:
\[
w_t = (1-B)(1-B^s)\ln\sigma_t
\]
where $B$ is the back-shift operator and $s=33$ is the seasonality.

The autocorrelation function of the differenced process $w_t$, as shown in
figure \ref{fig:nordea_15min_w_acf}, clearly points to a seasonal
moving-average model: There are only 4 non-zero autocorrelations in
the plot, located at lags 1, 32, 33, 34, respectively; furthermore,
the two at 32 and 34 are approximately equal. Thus we can write down
the model as
\begin{eqnarray}
  w_t &=& (1 - \theta B)(1 - \Theta B^s) y_t \label{eq:nordea_w}
\end{eqnarray}
where $\theta$ and $\Theta$ are parameters to be determined and $y_t$
is a noise process with constant variance $\sigma_y^2$ and mean
0. $y_t$ is often refered to as the residuals.
\begin{figure}[htb!]
  \centering
  \subfigure[ACF of $\ln\sigma_{t+h,h}$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_logvol_acf.pdf}
    \label{fig:nordea_15min_logvol_acf}
  }
  \subfigure[ACF of $w_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    235]{../pics/nordea_15min_w_acf.pdf}
    \label{fig:nordea_15min_w_acf}
  }
  \caption{\small \it Nordea 15min $\ln\sigma_{t+h, h}$ and $w_t$
    autocorrelations}
\end{figure}

The above seasonal moving average model has the following
autocovariance structure \cite{BoxJenkins94}:
\begin{eqnarray*}
  \gamma_0 &=& \sigma_y^2 (1 + \theta^2)(1 + \Theta^2) \\
  \gamma_1 &=& -\sigma_y^2\theta(1 + \Theta^2) \\
  \gamma_s &=& -\sigma_y^2\Theta(1 + \theta^2) \\
  \gamma_{s+1} &=& \gamma_{s-1}\;=\;\sigma_y^2\theta\Theta
\end{eqnarray*}
These equations together with the measured autocorrelations make
possible an initial estimate of the parameters $\theta$ and $\Theta$:
\begin{eqnarray*}
  {\rho_{s+1}/\rho_s} &=& {\gamma_{s+1}/\gamma_s} \;=\; -{\theta \over
    1 + \theta^2} \\
  {\rho_{s+1}/\rho_1} &=& {\gamma_{s+1}/\gamma_1} \;=\; -{\Theta \over
    1 + \Theta^2} \\
\end{eqnarray*}
Substituting in the measured values shown in table
\ref{tab:nordea_15min_w_acf},
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    $\rho_1$ & $\rho_{s-1}$ & $\rho_s$ & $\rho_{s+1}$ \\
    \hline
    -0.4703 &  0.2053 & -0.4564 &  0.2212 \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min $w_t$ autocorrelations}
  \label{tab:nordea_15min_w_acf}
\end{table}
we get
\begin{eqnarray*}
  \theta &=& 0.6890 \\
  \Theta &=& 0.6378
\end{eqnarray*}
Among the two roots of each of the 2nd order equations in the above,
we have chosen the one in the range $(-1, 1)$ so as to ensure
invertibility of the model \cite{BoxJenkins94}.

With an estimate of $\theta$ and $\Theta$, one can then infer the
noise process i.e. the residuals $y_t$:
\begin{equation}
  \label{eq:infer_y}
  y_t = w_t + \theta y_{t-1} + \Theta y_{t-s} - \theta \Theta y_{t-s-1}
\end{equation}
where we substitute $y_t\;(t \leq 0)$ with their unconditional
expectation 0.
% By inverting the seasonal moving average model of $w_t$ we can then
% infer the noise process $y_t$ \footnotemark, i.e. the residuals:
% \begin{eqnarray*}
%   y_t &=& {1 \over 1 - \theta B}{1 \over 1 - \Theta B^s} w_t \\
%   &=& \sum_{i=0}^{\infty}\sum_{j=0}^{\infty} \theta^i \Theta^j B^{i+j}
%   w_t \\
%   &=& w_t + \sum_{k=1}^{\infty} \sum_{j=0}^{\bot(k/s)} \theta^{k-sj} \Theta^j
%   w_{t-k}
% \end{eqnarray*}
% where $\bot(\cdot)$ denotes the floor function, i.e. the largest
% integer smaller than its argument.
% \footnotetext{
%   The autoregressive process resulting from the inversion is inifinite
%   in extent. Hence one must choose a cut-off value for k. For the
%   present case of Nordea 15min log-volatility, I choose the value
%   $k=sj$ for which $|\Theta|^j < 0.02$. This value of k, call it
%   $k_{\top}$, is around 550. To avoid data reduction of this amount,
%   we add to the beginning of $\{w_t\}_{t=1}^n$ $k_{\top}$ zeros
%   corresponding to t=0, -1, ..., -$k_{\top}$+1. Recall $\E(w_t) = \E(y_t
%   - \theta y_{t-1} - \Theta y_{t-s} + \theta \Theta y_{t-s-1}) = 0$.
% }

In order to forecst the $w_t$ process, and hence the return process
itself, we must also know the distribution of $y_t$. Moreover, to
properly estimate the parameters of the model in the sense of maximum
likelihood, we are also in need of the distribution of $y_t$.

If $y_t$ is indeed i.i.d random variables as we assumed, we only need
to work out its unconditional distribution. This is shown in figure
\ref{fig:nordea_15min_y_qq} and \ref{fig:nordea_15min_y_acf}.
\begin{figure}[htb!]
  \centering
  \subfigure[QQ-plot of $y_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    210]{../pics/nordea_15min_y_qq.pdf}
    \label{fig:nordea_15min_y_qq}
  }
  \subfigure[ACF of $y_t$]{
    \includegraphics[scale=0.4, clip=true, trim=95 230 112
    210]{../pics/nordea_15min_y_acf.pdf}
    \label{fig:nordea_15min_y_acf}
  }
  \caption{\small \it Nordea 15min $y_t$ QQ-plot and autocorrelations}
\end{figure}
In addition, the measured moments of $y_t$ are shown in table
\ref{tab:nordea_15min_y_moments}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|}
    \hline
    mean & variance & skewness & kurtosis \\
    \hline
    0.0012 & 0.0935 & 0.2988 & 6.8691 \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min $y_t$ moments}
  \label{tab:nordea_15min_y_moments}
\end{table}

If we ignore the weak autocorrelations in this crude estimate of
$y_t$ and assume $y_t$ to be i.i.d random variables governed
by a skewed and leptokertic distribution, the parameters of this
distribution together with $\theta$ and $\Theta$ can be determined via
maximum likelihood estimate (MLE).

It turns out that $y_t$ can be well described by a Johnson Su
distribution \cite{Shang2004}:
\[
  y_t = \xi + \lambda\sinh{z_t - \gamma \over \delta}
\]
where $\gamma, \delta, \lambda, \xi$ are parameters to be determined
and $z_t \sim N(0, 1)$. The goodness of fitting is demonstrated in
figure \ref{fig:nordea_15min_y_js_fit} by the
empirical cummulative distribution function in comparison to the
theoretical one.
\begin{figure}[htb!]
  %\vspace{-18mm}
  \centering
    \includegraphics[scale=0.5, clip=true, trim=92 229 116
    133]{../pics/nordea_15min_y_js_fit.pdf}
    \caption{\small \it Nordea 15min $y_t$ fitted to a Johnson Su
      distribution. Horizontal: values of $y_t$, denoted x; Vertical:
      $\ln\left(P(y_t < x)\right)$.}
    \label{fig:nordea_15min_y_js_fit}
\end{figure}

The first 4 moments of the Johnson Su distribution are expressible
in closed form in $\gamma, \delta, \lambda, \xi$ \cite{Shang2004}:
\begin{eqnarray*}
  w &=& \exp{1 \over \delta^2} \\
  \Omega &=& {\gamma \over \delta} \\
  \text{E}(y) &=& -w^{1/2} \lambda \sinh\Omega + \xi\\
  \text{std}(y) &=& \lambda \left[{1 \over 2}(w-1)(w\cosh 2\Omega +
    1)\right]^{1/2} \\
  \text{skewness}(y) &=& {
    \sqrt{(1/2)w(w-1)} [w(w+2)\sinh 3\Omega + 3\sinh\Omega]
    \over
    (w\cosh 2\Omega + 1)^{3/2}} \\
  \text{kurtosis}(y) &=& {
    w^2(w^4 + 2w^3 + 3w^2 - 3)\cosh 4\Omega + 4w^2 (w+2) \cosh 2\Omega
    + 3(2w+1) \over
    2(w\cosh 2\Omega + 1)^2 }
\end{eqnarray*}
By matching the theoretical expressions of the moments with their
measured values, and taking help from published tables
\cite{Johnson1965}, one can solve for the parameters $\gamma, \delta,
\lambda, \xi$.

Under the assumption of i.i.d Johnson Su distributed residuals, the
log-likelihood function of the parameters $\gamma, \delta$
conditional on the sample $w_t$ can be written as
\[
L(\theta, \Theta) = -{1 \over 2}\sum_{t=1}^n z_t^2 + n \ln{\delta
  \over \lambda \sqrt{2\pi}} - {1 \over 2}\sum_{t=1}^n \ln\left[
  1 + \left({y_t - \xi \over \lambda}\right)^2
\right]
\]
where $y_t$ are inferred from $w_t$ using eq.\ref{eq:infer_y}
and $z_t$ from $y_t$ using
\[
z_t = \delta \sinh^{-1}{y - \xi \over \lambda} + \gamma
\]
Note that $\gamma, \delta, \lambda, \xi$ are not really free
parameters but rather are implied by $\theta$ and $\Theta$: Once the
latter have been chosen and the corresponding $y_t$ inferred, the
former are determined by the moments of $y_t$.

The eventual MLE is done in Matlab with the ``active set''
algorithm. The initial as well as the final estimation results are
listed in table \ref{tab:nordea_15min_js_param}:
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|c|c|}
    \hline
    & $\gamma$ & $\delta$ & $\lambda$ & $\xi$ & $\theta$ & $\Theta$ \\
    \hline
    initial estimate & 0.1476 & 1.5121 & 0.3633 & 0.0454 & 0.6890 &
    0.6378 \\
    \hline
    MLE estimate & 0.1319 & 1.5266 & 0.3735 & 0.0410 & 0.6639 & 0.6025
    \\
    \hline
  \end{tabular}
  \caption{\small \it Nordea 15min estimation results}
  \label{tab:nordea_15min_js_param}
\end{table}

To verify that the log-volatility can in general be described by
ARIMA(p, d, q) models, we apply the above procedure to a number of
Swedish stocks whose data have been collected by the author with
home-developed software. In \S\ref{sec:nordea} and \S\ref{sec:volvo}
we model the log-volatilities of Nordea Bank 15-minute returns and
Volvo B 30-minute returns. Both return series span the period
2013/10/10 - 2014/04/04. They are computed using minute-by-minute
average prices, and their variances, i.e. volatilities squared, are
computed as the sum of squared higher-frequency returns, which are in
turn computed using second-by-second prices. Whenever a needed average
price cannot be computed due to the lack of price observations in the
relevent time interval, the value of the previous average price is
assumed.

After modeling, we perform 1-step-ahead forecast using the estimated
models, and compare the predicted log-volatilities with the actual
data.

\subsection{Nordea Bank Log-Volatility}\label{sec:nordea}
In this section we model the log-volatility of {\it Nordea Bank}
15-minute returns during the period 2013/10/10 - 2014/04/04, which
contains 3879 data points. The squared volatilities, i.e. the
variances, are computed as the sum of squared 30-second returns. Among
the 3879 data points, we use the first 80\% for model estimation and
the last 20\%, i.e. 776 data points for forecast verification.

This Nordea series has a auto-correlation structure similar to its
predecessor that has been studied in previous sections. We show it in
figure \ref{fig:nordea2_w_acf}. The reader may want to compare it with
figure \ref{fig:nordea_15min_w_acf}.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.4, clip=true, trim=95 230 112
  235]{../pics/nordea2_w_acf.pdf}
  \caption{\small \it Auto-correlation function of $w_t$}
  \label{fig:nordea2_w_acf}
\end{figure}
The significant auto-correlations and their respective lags are listed
in table \ref{tab:nordea2_w_acf}.
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    lag & 1 & 32 & 33 & 34 \\
    \hline
    ACF &   -0.4805 & 0.2861 &  -0.5034 & 0.2093 \\
    \hline
  \end{tabular}
  \caption{\small \it Significant auto-correlations of $w_t$}
  \label{tab:nordea2_w_acf}
\end{table}

Apparently this series also has the seasonality $s = 33$, but judging
from the relatively big differences between auto-correlations at lags
32 and 34, we don't expect the moving average process to be in the
multiplicative form of equation \ref{eq:nordea_w}. Instead, we model
$\ln\sigma_t$ has
\begin{eqnarray}
  (1 - B)(1 - B^s) \ln\sigma_t &=& (1 - \theta_1 B - \theta_s B^s -
  \theta_{s+1} B^{s+1} ) y_t \label{eq:nordea2_v}
\end{eqnarray}
The parameters are etimated to be
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    parameter & $\theta_1$ & $\theta_{33}$ & $\theta_{34}$ & $\var(y_t)$ \\
    \hline
    value & 0.8339 & 0.8448 & -0.7067 & 0.10071 \\
    \hline
  \end{tabular}
  \caption{\small \it: Parameter values of $\ln \sigma_t$}
\end{table}

The residuals $y_t$ do not have a Gaussian distribution, as confirmed
by the probability plot shown on the left of figure
\ref{fig:nordea2_y}. The non-normality, however, cannot be accounted
by the sine hypobolic transformation of Johnson Su, because of the
negative skewness of -1.2669 found in table
\ref{tab:nordea2_y_mmt}. How the non-normality can be accomodated is
quite another discussion and we do not pursue it here.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.4, clip=true, trim=20 220 30
  115]{../pics/nordea2_y.pdf}
  \caption{\small \it The residuals' distribution and auto-correlations. Left:
    cummulative distribution function (CDF) of the residuals $y_t$; Right:
    auto-correlation function of $y_t$.}
  \label{fig:nordea2_y}
\end{figure}

\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Moment & mean & variance & skewness & kurtosis \\
    \hline
    value & -0.0022 & 0.1007 & -1.2669 & 15.5965 \\
    \hline
  \end{tabular}
  \caption{\small \it: Moments of the residuals $y_t$}
  \label{tab:nordea2_y_mmt}
\end{table}

The auto-correlations of the residuals $y_t$ are insignificant, as
shown on the right of figure \ref{fig:nordea2_y}. Though the Ljung-Box Q
test rejects the null hypothesis of no-autocorrelation in the
residuals, the weak auto-correlations, if any at all,  are not worth
further complicating the model.

Forecast of $\ln\sigma_t$ is done by reverting equation
\ref{eq:nordea2_v} and write
\begin{eqnarray*}
  \ln\sigma_t &=& \ln\sigma_{t-1} + \ln\sigma_{t-s} -
  \ln\sigma_{t-s-1} + (1 - \theta_1 B - \theta_s B^s + \theta_{s+1}
  B^{s+1} ) y_t \\
  \E(\ln\sigma_t | \mathcal{F}_{t-1}) &=& \ln\sigma_{t-1} + \ln\sigma_{t-s} -
  \ln\sigma_{t-s-1} - \theta_1 y_{t-1} - \theta_s y_{t-s} +
  \theta_{s+1} y_{t-s-1}
\end{eqnarray*}

Thus the forecast, namely $\E(\ln\sigma_t | \mathcal{F}_{t-1})$, is
found in terms of previous values in the same series and of previous
residuals.

To compare the performance, we also estimate a GARCH model with
Gaussian conditional distribution for the Nordea Bank 15-minute
return. The procedure is standard and we only list the results here
\begin{table}[htb!]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Parameter & $\alpha_0$ & $\alpha_1$ & $\alpha_5$ & $\beta_1$ \\
    \hline
    Value & $7.3686 \times 10^{-7}$ & 0.154131 & 0.0360 & 0.491832 \\
    \hline
  \end{tabular}
  \caption{\small \it Parameters of GARCH(1, 5) model of Nordea
    Bank 15-minute returns}
  \label{tab:nordea2_garch}
\end{table}

Forecasting using both models, we get two series of one-step-ahead
forecast, $\ln \sigma^{\text{ARIMA}}_t$ and $\ln \sigma^G_t$,
respectively. We denote the empirical log-volatilities as $\ln
\hat{\sigma}_t$. Then the accuracy of the two models can be compared
by comparing $\ln \sigma^{\text{ARIMA}}_t - \ln \hat{\sigma}_t$ and
$\ln \sigma^G_t - \ln \hat{\sigma}_t$. As a reference, we also include
in the comparison the naive approach of taking the sample mean of
$\ln \sigma_t$ as the forecast.

>>>>

\chapter{Unconditional Distribution Functions of Stochastic
  Log-Volatility Models}
\label{chp:SLV_unconditional}
In this chapter we study the unconditional distribution functions of
Stochastic Log-Volatility models specified as equation
\ref{eq:SLV_spec}. As is discussed in section \ref{sec:SLV_model},
$\ln\sigma_t$ can be described by an ARMA or ARIMA \footnote{ARIMA ---
  Integrated Autoregressive Moving Average} model, possibly with
seasonal components. Here we note that all these models can be
re-written as a moving average model, which is infinite in extent if
autoregressive components are present:
\begin{eqnarray*}
  \ln \sigma_t &=& y_t + \sum_{n=1}^\infty c_n y_{t-n} + \text{Const.}
\end{eqnarray*}
Since the $y_t$ are independent and identically distributed,
\begin{eqnarray*}
  y_t + \sum_{n=1}^\infty c_n y_{t-n}  
\end{eqnarray*}
has Gaussian distribution by the central limit theorem, on condition
that $y_t$ for all $t$ have finite second moment --- this is what we
assume in the rest of this section. It follows from the above equation
that the unconditional distribution of $\ln \sigma_t$ is the same as the
distribution of $\bar{v} + v$ where $v \sim N(0, \sigma)$ and
$\bar{v}$, $\sigma$ are constants. Now we can state that the
unconditional distribution of the returns $r_t$
\begin{eqnarray*}
  r_t &=& \mu + \sigma_t b_t\\
  &=& \mu + \exp\left(
    y_t + \sum_{n=1}^\infty c_n y_{t-n} + \text{Const.}
  \right)
\end{eqnarray*}
 is the same as
\begin{equation}  \label{eq:UnconditionalPdf}
  \begin{aligned}
    r &= \mu + e^{\bar{v} + v} b \\
  \end{aligned}
\end{equation}
where $b \sim N(0, 1)$. For convenience, let $r' = e^v b$

In section \ref{sec:SLV_Symmetric} we first study the model in the
relatively simple case when $v$ and $b$ are uncorrelated and $\mu
= 0$. If this simplified version proves inadequate, one may resort to
the general model studied in section \ref{sec:SLV_Asymmetric}.

\section{The Simplified model}\label{sec:SLV_Symmetric}
In the following we derive the unconditional PDF of $r'$, denoted
$f_{r'}(x)$. Then the PDF of $r$ is $e^{-\bar{v}}f_{r'}(e^{-\bar{v}}x)$.
\begin{eqnarray*}
  P(r' < x) &=& P(b < xe^{-v}) \\
  f_{r'}(x) &=& f_b(xe^{-v}) e^{-v}
\end{eqnarray*}
Averaging over all $v$, we get
\begin{equation}\label{eq:UncondPDFSymmetric}
  \begin{aligned}
    f_{r'}(x) =& \int_{-\infty}^{\infty} dv (2\pi\sigma^2)^{-1/2}
    e^{-v^2/2\sigma^2}(2\pi)^{-1/2} \exp(-x^2e^{-2v}/2) e^{-v} \\
    =& {1 \over 2\pi\sigma} \int_{-\infty}^{\infty} dv
    \exp\left(
      -{1 \over 2\sigma^2} v^2 - v -{1 \over 2} x^2 e^{-2v}
    \right)
  \end{aligned}
  \end{equation}
The last part of the integrand, $e^{-x^2 e^{-2v} / 2}$, is plotted in
figure \ref{fig:DoubleExp}.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5, clip=true, trim=85 252 100
  231]{../pics/DoubleExp.pdf}
  \caption{\small \it Plot of $\exp(-{1 \over 2} x^2 e^{-2v})$}
  \label{fig:DoubleExp}
\end{figure}
Therefore we make the following approximation:
\[
\exp\left(-{1 \over 2} x^2 e^{-2v}\right) \approx \left\{
  \begin{array}{lr}
    0 & \text{if } v < \ln|x| -{1 \over 2} \ln(2\ln 2) \\
    1 & \text{otherwise}
  \end{array}
\right.
\]
Here we note that $\exp\left(-{1 \over 2} x^2 e^{-2v}\right) = 1/2$ at
$v = \ln|x| -{1 \over 2} \ln(2\ln 2)$.
